---
title: "CleaningMortality"
author: "Ainsley Lotito"
date: "2025-11-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
library(readr)  
library(dplyr)
library(tidyverse)
library(forcats)
library(lubridate)
```


## Mortality Data

We will start by cleaning the mortality dataset from the national park service. 
Let's start by reading in the data and looking at the columns. 

```{r}
mortality <- read_csv("../data/NPS-Mortality-Data-CY2007-to-CY2024-Released-August-2024.csv", col_names = TRUE)
head(mortality) 
dim(mortality)
``` 

Mortality read in correctly, we have 9 variables and 4,635 observations. 

We need to reclassify some variables, rename them, and reclassify not reported information as NA. 
```{r} 
mortality <- mortality %>% 
  mutate(incident_date = mdy(`Incident Date`),
         year = year(incident_date)) %>% #creating year variable to aid with merging with other df
  rename(park = `Park Name`,
         death_cause = `Cause of Death`,
         death_group = 4,
         age_range = `Age Range`) %>% 
  select(-1,) 
colnames(mortality) <- tolower(colnames(mortality))
head(mortality)

#changing "Not Reported" to be NA
mortality <- mortality %>% 
  mutate(across(c(intent, death_group, death_cause),
                ~ na_if(., "Undetermined")),
         across(c(age_range, activity),
                 ~ na_if(., "Not Reported")))
head(mortality)


```


We are going to subset our data to include years 2007-2016 to be consistent with the other dataset. 