---
title: "CleaningMortality"
author: "Ainsley Lotito"
date: "2025-11-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
library(readr)  
library(dplyr)
library(tidyverse)
library(forcats)
library(lubridate) 
library(naniar)
```


## Mortality Data

We will start by cleaning the mortality dataset from the national park service. 
Let's start by reading in the data and looking at the columns. 

```{r}
mortality <- read_csv("../data/NPS-Mortality-Data-CY2007-to-CY2024-Released-August-2024.csv", col_names = TRUE)
head(mortality) 
dim(mortality)
``` 

Mortality read in correctly, we have 9 variables and 4,635 observations. 

We need to reclassify some variables, rename them, and reclassify not reported information as NA. 
```{r} 
mortality <- mortality %>% 
  mutate(incident_date = mdy(`Incident Date`),
         year = year(incident_date)) %>% #creating year variable to aid with merging with other df
  rename(park = `Park Name`,
         death_cause = `Cause of Death`,
         death_group = 4,
         age_range = `Age Range`) %>% 
  select(-1,) 
colnames(mortality) <- tolower(colnames(mortality))
head(mortality)
``` 

We are going to explore missing data by reclassifying "Undetermined" and "Not reported" data in a new dataframe and look through some patterns to see if specific causes of death have higher proportions of missing data. 
```{r}
#changing "Not Reported" to be NA in different dataset
mortality_miss <- mortality %>% 
  mutate(across(c(intent, death_group, death_cause),
                ~ na_if(., "Undetermined")),
         across(c(age_range, activity),
                ~ na_if(., "Not Reported"))) 
gg_miss_var(mortality_miss)
gg_miss_upset(mortality_miss) 

mortality_miss %>% 
  mutate(missing = is.na(intent) & is.na(age_range) & is.na(activity)) %>% 
  group_by(death_cause, missing) %>% 
  count()

mortality_miss %>% 
  mutate(missing =  is.na(age_range)) %>% 
  group_by(death_cause, missing) %>% 
  count()

#add on later to summarize by proportions and which have highest mortality counts
```


We are going to subset our data to include years 2007-2016 to be consistent with the other dataset.  
```{r} 
#seeing how many obs we are cutting out

 mortality %>% 
  filter(year %in% c(2020:2024)) %>% 
   nrow()
#we are cutting out 2017 ()

mortality <- mortality %>% 
  filter(year %in% c(2007:2019))
dim(mortality) 
unique(mortality$year) 
```

We now have 3,262 observations (it removed 1,373 obsevations)  

### Saving cleaned dataset 
We are going to save the cleaned mortality df to our data folder so it can be joined with the National Park dataframe for future analysis. 

```{r}
write.csv(mortality, "../data/mortality_clean.csv", row.names = TRUE)
```

# Exploratory analysis 

We are interested in looking at patterns of mortality in our data before we join the datasets to think through maps that would be interesting. 
First, we are going to plan out a chloropleth map that organizes mortality counts in the parks over the whole period (2007-2019) and by each year individually. 

We will start by analyzing this with the summarize function first to look at which parks have the highest mortality counts over each year. 
```{r}
#raw count
mortality %>% 
  group_by(park) %>%  
  summarize(mortality_count = n()) %>% 
  arrange(desc(mortality_count)) 

#top 3 by year 
mortality %>% 
  group_by(park, year) %>%  
  summarize(mortality_count = n(), .groups = "drop") %>% 
  group_by(year) %>% 
  slice_max(mortality_count, n = 3)

``` 

We also want to look into the causes of death and how those contribute to mortality counts. 
```{r}
#raw count
mortality %>% 
  group_by(death_group) %>%  
  summarize(mortality_count = n()) %>% 
  arrange(desc(mortality_count)) 

#top 3 by year 
mortality %>% 
  group_by(death_group, year) %>%  
  summarize(mortality_count = n(), .groups = "drop") %>% 
  group_by(year) %>% 
  slice_max(mortality_count, n = 3) 

#let's do it again by age group 
#raw count
mortality %>% 
  group_by(death_group, age_range) %>%  
  summarize(mortality_count = n()) %>% 
  arrange(desc(mortality_count)) 

#top 3 by year 
mortality %>% 
  group_by(death_group, age_range, year) %>%  
  summarize(mortality_count = n(), .groups = "drop") %>% 
  group_by(year) %>% 
  slice_max(mortality_count, n = 3)  

#top 3 by park 
mortality %>% 
  group_by(death_group, age_range, park) %>%  
  summarize(mortality_count = n(), .groups = "drop") %>% 
  group_by(death_group,age_range) %>% 
  slice_max(mortality_count, n = 1) 
``` 

Let's explore some leaflet bar charts with different information on cause of death.  

```{r bar_chart}
library(plotly)

#starting simple plot  
cod_count <- mortality %>% 
  group_by(death_group) %>%  
  summarize(mortality_count = n(), .groups = "drop")

cod_plot <-plot_ly(
  data = cod_count,
  x = ~death_group,
  y = ~mortality_count, 
  name = "Death Counts by Cause of Death", 
  type = "bar"
) 
cod_plot <- cod_plot %>% layout(yaxis = list(title = 'Mortality Count'), barmode = 'group') %>% layout(xaxis = list(title = 'Cause of Death'), barmode = 'group')

cod_plot
``` 
We can probably remove "legal intervention", since there are only 2 deaths over the entire period. These deaths are due to people being killed or dying as a result of injuries from a law enforcement officer.