---
title: "Exploring deaths in National Parks in the U.S."
author: "Jackson DeCook, Neda Safari, Kalex Robedo, Ainsley Lotito"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    social: menu
    source_code: embed 
    runtime: shiny
---

```{r setup, include=FALSE}
library(highcharter)
library(dplyr)
library(viridisLite)
library(arules)
library(forecast)
library(treemap)
library(flexdashboard) 
library(shiny)
library(plotly)


thm <- 
  hc_theme(
    colors = c("#1a6ecc", "#434348", "#90ed7d"),
    chart = list(
      backgroundColor = "transparent",
      style = list(fontFamily = "Source Sans Pro")
    ),
    xAxis = list(
      gridLineWidth = 1
    )
  ) 

```

## Column {data-width="350"}

### About This Dashboard

The source of this data was the Tidy Tuesday data set involving national park visitation rates and boundaries. National Park Service (NPS) data on deaths is also included to help answer questions about national park areas with many deaths, common causes of death, and age groups occur most frequently. 

The *Interactive Mortality Map* displays the total death counts at all national parks in the United States. Scroll to zoom in and hover over individual parks to find their attributed death count. The *MNational Park Death Trends Search Tool* also displays individual park death counts, as well as the common age ranges and causes of death. Finally, users can select an *Deaths by Cause and Age Range* from the cause of death drop-down menu to explore how deaths are distributed across age ranges.

Interestingly, we found that of drowning deaths in national parks, many were within the 15-24 age range instead of the expected 0-14 age range. This is likely due to the heightened awareness and regulations surrounding drowning safety in children, while these are softened in teenagers and young adults. Also, other factors such as alcohol could affect the increase in drowning deaths in this age group.


## Column {data-width="600"}

### Interactive Mortality Map

```{r}
library(tidyverse)
library(highcharter)
park_map_data <- read.csv("../data/park_map_data.csv")
# Create the interactive map
# hcmap() creates a base map of the United States
# "countries/us/us-all" specifies we want a US map
# nullColor sets the color for states/areas without data
# borderWidth sets how thick the state borders are
# showInLegend = FALSE hides the base map from the legend
hcmap("countries/us/us-all", 
      nullColor = "#d3d3d3",
      borderWidth = 0.5,
      showInLegend = FALSE) %>%
  # hc_add_series adds a new layer to the map - in this case, bubbles
  hc_add_series(
    data = park_map_data,  # The data frame with park locations and death counts
    type = "mapbubble",  # Create bubbles on the map
    name = "Deaths",  # Legend label
    minSize = 5,  # Smallest bubble size (in pixels)
    maxSize = 50,  # Largest bubble size (in pixels)
    color = "#FF0000",  # Bubble color (red)
    dataLabels = list(enabled = FALSE)  # Don't show labels on the bubbles themselves
  ) %>%
  # hc_tooltip configures what appears when you hover over a bubble
  hc_tooltip(
    useHTML = TRUE,  # Allow HTML formatting in the tooltip
    headerFormat = '',  # No header in the tooltip
    # pointFormat defines what text shows: park name and death count
    pointFormat = "<b>{point.park}</b><br/>Total Deaths: {point.total_deaths}"
  ) %>%
  # hc_mapNavigation adds zoom and pan controls to the map
  hc_mapNavigation(enabled = TRUE) %>%
  # hc_title adds a title at the top of the map
  hc_title(text = "National Park Mortality Counts")
```

### National Park Death Trends Search Tool

```{r}
library(tidyverse)
library(DT)
MortalityData<- read.csv("../data/mortality_clean.csv")
mortalityCount<- MortalityData%>%
  group_by(park, death_cause, age_range) %>%
  summarize(death_count=n())%>%
  ungroup()%>%
  arrange(desc(death_count))

datatable(mortalityCount, options = list(pageLength = 5))

```

## Column {.tabset data-width="400"}

### Deaths by Cause and Age Range

```{r, fig.keep='none'}
mortality <- read.csv("../data/mortality_clean.csv")

# summarizing cod by age
cod_by_age <- mortality %>% 
  group_by(cod, age_range) %>%  
  summarize(mortality_count = n(), .groups = "drop")

cod_order <- cod_by_age %>%
  group_by(cod) %>%
  summarize(total_mort = sum(mortality_count)) %>%
  arrange(desc(total_mort))

cod_by_age$cod <- factor(cod_by_age$cod, levels = cod_order$cod)

#UI Function
ui <- fluidPage( #creates a fluidPage
  sidebarLayout( #creates sidebar on the left
    sidebarPanel( #what's in the sidebar
      selectInput("cod_choice", "Select Cause of Death:", 
                  choices = levels(cod_by_age$cod)) #cod_choic is input ID, "Select Cause of Death:" is label shown, choices -- populate the dropdown with faactor levels from the cod column
    ),
    mainPanel( #starting the main panel
      plotlyOutput("codPlot") #placeholder for a plotly plot
    )
  )
)

server <- function(input, output) { #server- takes inputs, produce output
  output$codPlot <- renderPlotly({ #create reacative plotly output named "codPlot". everything INSIDE CURLY runs when the user changes the dropdown
    
    filtered <- cod_by_age %>% 
      filter(cod == input$cod_choice) #only include rows where cod= user's selection
    
    p <- ggplot(filtered, aes(x = age_range, y = mortality_count)) +
      geom_col() +
      labs(
        title = paste("Deaths by Age for:", input$cod_choice),
        x = "Victim Age Group",
        y = "Death Count"
      ) +
      theme_classic()
    #ggplot of the filtered data 
    ggplotly(p) #turns into plotly
  })
}

shinyApp(ui, server) #ui=page layout for shiny, server= plotly 
#runs the App using the UI and server we just defined

```

